---
title: "2023 NCAA March Madness Predictor"
output: html_document
date: "2024-03-14"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# put your library statements here
library(tidyverse)
library(caret)
library(ROCR)
library(dplyr)

# import the data
ncaa_data <- read_csv("https://raw.githubusercontent.com/gmtanner-cord/DATA318/master/Original%20Data/ncaa_data_2001_2023.csv")
ncaa_data <- ncaa_data %>% mutate(winner = factor(winner, levels = c("team1","team2")))
```

Model A predicts the point differential in college basketball games based on four key predictors: seed difference, win percentage difference, Simple Ratings System difference, and points per game difference between opposing teams. The model's performance is assessed using R-squared, p-values, and RMSE. Through visualizations, the impact of each predictor on the predicted point differential is examined, allowing for insights into how various factors influence game outcomes.

```{r}
ncaa_data <- ncaa_data %>%
  mutate(seed_diff = seed.1 - seed.2,
         win_pct_diff = win_pct.1 - win_pct.2,
         srs_diff = srs.1 - srs.2,
         pts_pg_diff = pts_pg.1 - pts_pg.2)

model_reg <- lm(point_diff ~ seed_diff + win_pct_diff + srs_diff + pts_pg_diff, data = ncaa_data)
summary(model_reg)

grid <- expand.grid(seed_diff = seq(-15, 15, 1),
                    win_pct_diff = 0,
                    srs_diff = mean(ncaa_data$srs_diff),
                    pts_pg_diff = mean(ncaa_data$pts_pg_diff))

yhat <- predict(model_reg, newdata = grid)
ggplot() +
  geom_jitter(data = ncaa_data, aes(x = seed_diff, y = point_diff), alpha = 0.3) +
  geom_line(data = grid, aes(x = seed_diff, y = yhat))

grid <- expand.grid(seed_diff = 0,
                    win_pct_diff = quantile(ncaa_data$win_pct_diff),
                    srs_diff = mean(ncaa_data$srs_diff),
                    pts_pg_diff = mean(ncaa_data$pts_pg_diff))

yhat <- predict(model_reg, newdata = grid)
ggplot() +
  geom_point(data = ncaa_data, aes(x = win_pct_diff, y = point_diff), alpha = 0.3) +
  geom_line(data = grid, aes(x = win_pct_diff, y = yhat))

grid <- expand.grid(seed_diff = 0,
                    win_pct_diff = mean(ncaa_data$win_pct_diff),
                    srs_diff = quantile(ncaa_data$srs_diff),
                    pts_pg_diff = mean(ncaa_data$pts_pg_diff))

yhat <- predict(model_reg, newdata = grid)
ggplot() +
  geom_point(data = ncaa_data, aes(x = srs_diff, y = point_diff), alpha = 0.3) +
  geom_line(data = grid, aes(x = srs_diff, y = yhat))

grid <- expand.grid(seed_diff = 0,
                    win_pct_diff = mean(ncaa_data$win_pct_diff),
                    srs_diff = mean(ncaa_data$srs_diff),
                    pts_pg_diff = quantile(ncaa_data$pts_pg_diff))

yhat <- predict(model_reg, newdata = grid)
ggplot() +
  geom_point(data = ncaa_data, aes(x = pts_pg_diff, y = point_diff), alpha = 0.3) +
  geom_line(data = grid, aes(x = pts_pg_diff, y = yhat))
```

Model B is a logistic regression that predicts the winner of college basketball games based on factors like the difference in seed and win percentage between teams. The model's performance is assessed using measures like p-values, ROC curve, AUC, kappa statistic, and accuracy. The ROC curve shows how well the model distinguishes between true positives and false positives. AUC tells us how effective the model is at this task. Kappa indicates how well the model's predictions match actual outcomes beyond random chance, while accuracy simply shows the overall correctness of the predictions.

```{r}
model_class <- glm(winner ~ seed_diff + win_pct_diff, data = ncaa_data, family = "binomial")

summary(model_class)

pvalues_class <- summary(model_class)$coefficients[, 4]

predicted_winner_prob <- predict(model_class, type = "response")

predicted_winner <- ifelse(predicted_winner_prob > 0.5, "team1", "team2")

accuracy <- mean(predicted_winner == ncaa_data$winner)

ROCRpred <- prediction(predicted_winner_prob, ncaa_data$winner)
ROCRperf <- performance(ROCRpred, "tpr", "fpr")
plot(ROCRperf)

auc <- performance(ROCRpred, "auc")
auc <- auc@y.values[[1]]

predicted_winner <- factor(predicted_winner, levels = c("team1", "team2"))

kappa <- confusionMatrix(predicted_winner, ncaa_data$winner)$overall["Kappa"]

pvalues_class
cat("Accuracy:", accuracy, "\n")
cat("AUC:", auc, "\n")
cat("Kappa:", kappa, "\n")
```
