---
title: "2023 NCAA March Madness Predictor"
output: html_document
date: "2024-03-14"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# put your library statements here
library(tidyverse)
library(caret)
library(ROCR)
library(dplyr)

# import the data
ncaa_data <- read_csv("https://raw.githubusercontent.com/gmtanner-cord/DATA318/master/Original%20Data/ncaa_data_2001_2023.csv")
ncaa_data <- ncaa_data %>% mutate(winner = factor(winner, levels = c("team1","team2")))
```

Model A predicts the point differential in college basketball games based on four key predictors: seed difference, win percentage difference, Simple Ratings System difference, and points per game difference between opposing teams. The model's performance is assessed using R-squared, p-values, and RMSE. Through visualizations, the impact of each predictor on the predicted point differential is examined, allowing for insights into how various factors influence game outcomes.

```{r}
ncaa_data <- ncaa_data %>%
  mutate(seed_diff = seed.1 - seed.2,
         win_pct_diff = win_pct.1 - win_pct.2,
         srs_diff = srs.1 - srs.2,
         pts_pg_diff = pts_pg.1 - pts_pg.2)

model_reg <- lm(point_diff ~ seed_diff + win_pct_diff + srs_diff + pts_pg_diff, data = ncaa_data)
summary(model_reg)

grid <- expand.grid(seed_diff = seq(-15, 15, 1),
                    win_pct_diff = 0,
                    srs_diff = mean(ncaa_data$srs_diff),
                    pts_pg_diff = mean(ncaa_data$pts_pg_diff))

yhat <- predict(model_reg, newdata = grid)
ggplot() +
  geom_jitter(data = ncaa_data, aes(x = seed_diff, y = point_diff), alpha = 0.3) +
  geom_line(data = grid, aes(x = seed_diff, y = yhat))

grid <- expand.grid(seed_diff = 0,
                    win_pct_diff = quantile(ncaa_data$win_pct_diff),
                    srs_diff = mean(ncaa_data$srs_diff),
                    pts_pg_diff = mean(ncaa_data$pts_pg_diff))

yhat <- predict(model_reg, newdata = grid)
ggplot() +
  geom_point(data = ncaa_data, aes(x = win_pct_diff, y = point_diff), alpha = 0.3) +
  geom_line(data = grid, aes(x = win_pct_diff, y = yhat))

grid <- expand.grid(seed_diff = 0,
                    win_pct_diff = mean(ncaa_data$win_pct_diff),
                    srs_diff = quantile(ncaa_data$srs_diff),
                    pts_pg_diff = mean(ncaa_data$pts_pg_diff))

yhat <- predict(model_reg, newdata = grid)
ggplot() +
  geom_point(data = ncaa_data, aes(x = srs_diff, y = point_diff), alpha = 0.3) +
  geom_line(data = grid, aes(x = srs_diff, y = yhat))

grid <- expand.grid(seed_diff = 0,
                    win_pct_diff = mean(ncaa_data$win_pct_diff),
                    srs_diff = mean(ncaa_data$srs_diff),
                    pts_pg_diff = quantile(ncaa_data$pts_pg_diff))

yhat <- predict(model_reg, newdata = grid)
ggplot() +
  geom_point(data = ncaa_data, aes(x = pts_pg_diff, y = point_diff), alpha = 0.3) +
  geom_line(data = grid, aes(x = pts_pg_diff, y = yhat))
```
